\emph{icepack}: a new glacier flow modeling package in Python, version 1.0

### 2021-03-31 reviews/response-2.tex

It is possible to ruin the performance of the application by accidentally hard-coding a floating point value into a symbolic expression of a PDE and then changing that value in a loop.
For example you could easily make this mistake if you were doing adaptive timestepping on an evolutionary problem discretized via the method of lines.
The remedy is to wrap this value in a \texttt{firedrake.Constant} object.
But this kind of performance regression is easily caught using \texttt{htop} and are more a result of programmer error.

For the problems that we have used icepack for so far, we have focused more on individual glaciers or catchments, and thus performance has been a secondary concern.
We aim to move towards larger continental-scale problems in the future.
The rate-limiting factor there is more our ability to find the right incantation of PETSc solver options and preconditioners than it is any inherent limitation in our tools.

\item \textcolor{blue}{I was a little bit confused about the reference of the FO Stokes-based model in this paper as a ``hybrid
model''. I see that it is hybrid in the sense that you have a different discretization in the horizontal and
vertical direction, but there are also hybrid ice models that use different PDEs in different domains,
e.g., the ISCAL model of Ahlkrona et al. (https://doi.org/10.1016/j.quascirev.2016.01.032).
Is the term ``hybrid model'' a common name for the approach in Section 2.2.3? Perhaps it is and I
am not aware of it.
Does the hybrid model described in the paper have the same applicability as say the First Order
Stokes model? Can it be used for both Greenland and Antarctica at continental scales?}

### 2021-03-31 reviews/response-3.tex

\section*{Line-by-line comments}

\textcolor{blue}{13, AC:  Presumably all software packages can be effective in the hands of experts, and here it is suggested that Icepack stands-out as better for non-experts.  My question is whether it will be effective for non-experts interested in announced goals 1 and 4 (among those listed on lines 15--18).  This is hard to believe given how Icepack seems to be designed around, and the manuscript focused on, goals 2 and 3.  Consider the reader who wants to simulate glacier extent for some years into the future for a mountain glacier or Greenland, who can supply a simulated climate (atmosphere/ocean), a bed topography, and a current geometry in some data files.  Does this paper convince me that they need *less* expert knowledge to use Icepack for that purpose than other existing ice sheet models?  Not yet.  Of course a full usage answer occurs in online tutorials and examples, not just the manuscript.  Nonetheless the absence of attention to transport equation boundary conditions, and to how mass/energy surface inputs are handled, gives me the impression that such a reader is on a long co-development path with the Icepack authors, requiring the development of much expert knowledge before first useful results.  Said a different way, expert knowledge is required for any software that does not have an aggressive scheme for putting reasonable defaults into the hands of novices.  To paraphrase a recent Firedrake paper [Farrell et al 2020], it is a mechanism-vs-policy concern.  Icepack has a library of mechanisms, but an absence of apparent policy means expert knowledge is needed to recover usability for real-science applications.  I rolled my eyes at the implications of the sentence on line 13, and these concerns remained after reading the whole paper.}

This depends on what kind of expertise you mean.
We believe that more work can be done to relieve glaciologists \emph{who are not experts in scientific computing} from having to understand, e.g., the details of what preconditioners are used in solving linear systems, what globalization strategy is used in a Newton solver, etc.
By way of an ``aggressive scheme for putting reasonable defaults into the hands of novices'', we describe this policy at the end of section 6, beginning in line 695.
We do not claim to have reduced the need for expertise in glaciology itself, i.e. understanding mass and energy balance, etc.
We also consider having some understanding of the different boundary conditions that might apply to the system to be part of this knowledge.
The text has been expanded to describe these boundary conditions in more detail (see comments that follow).
Granted, (1) there is more demonstration work that we need to do in order to cover use cases like modeling a small mountain glacier and (2) there are technical hurdles we still need to overcome to do a better job handling the margins.
Very likely your recent work on variational inequalities will factor into improvement on the second front.
Addressing these concerns within this manuscript would result in a dramatic increase in scope.

@@

\textcolor{blue}{21--39:  I also think Python+Firedrake+PETSc is the most promising environment to build a new ice flow simulation library/model.  I'm on board!}

\textcolor{blue}{44--46, B:  Let me vote to *not* maintain the "diagnostic"/"prognostic" linguistic tribalism.  The world calls these equations "conservation of momentum" and "conservation of mass".  (Indeed one should remind the reader that the latter describes thickness evolution because of how glacier models normally parameterize fluid geometry.)  An "also known as the 'prognostic equation' [cite]" is appropriate, but another paper using this tribal language will cause yet more students to need to unlearn silly language in order to read the mainstream fluids and numerical PDE literature.}

I agree with you that this isn't great terminology.
But I didn't think it was a fight I could win so I went with the rest of the glaciological literature and used these names.
By this point the nomenclature is used in so many places throughout the package that switching would be a breaking change for existing users.

### 2021-04-01 icepack.tex
More responses to reviewer 3

There are four \emph{diagnostic} or \emph{momentum transport} models implemented in icepack.
For each of the diagnostic models, we use a formulation of the physics based on a \emph{minimization} or \emph{action principle} \citep{dukowicz2010consistent}.
Action principles are completely equivalent to the usual weak form of a partial differential equation but have certain numerical advantages, as described below.

@@

The action principle is especially useful for designing a robust numerical solver.
For viscous flow problems near to steady-state, the action is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
Convexity implies that the action functional has a unique minimizer and that, with an appropriate line search strategy, Newton's method will converge from any initial guess.
Minimizing a convex action functional is vastly more convenient numerically than solving a general nonlinear equation while having no additional computational cost.
Both formulations are mathematically equivalent.

@@

The shallow ice model can be derived from the Blatter-Pattyn approximation by assuming ice flow is dominated by bed-parallel shear and that surface and basal slopes are small, i.e. $\partial u/\partial x \ll \partial u/\partial z$.
The result is a 2D system of equations for depth-averaged horizontal velocity.
The class in icepack that represents this physics model is called \texttt{ShallowIce}.
The terms in the action functional are:
\begin{align}
    \text{mass} & =\frac{1}{2}\int_\Omega u\cdot u\,dx\\\
    \text{gravity} & = \int_\Omega\frac{2A(\rho_I g)^n}{n+2} h^{n + 1}|\nabla s|^{n - 1}\nabla s\cdot u\,dx\\
    \text{penalty} & = \frac{1}{2}\int_\Omega \ell^2\nabla u\cdot \nabla u\,dx
\end{align}
and the action functional is
\begin{equation}
    J = \text{mass} + \text{gravity} + \text{penalty}.
\end{equation}
The default value of the length scale $\ell$ in the penalty term is defined as:
\begin{equation}
    \ell = 2\max\{\text{cell diameter}, 5h\}
\end{equation}
but users can adjust this to the value of their choice.
This penalty term smooths over numerical artifacts, especially near the ice margins and termini.
In these regions the shallow ice approximation is less applicable, so the error in solving a different set of equations is small compared to the inherent modeling error in using these equations in the first place.

@@

We verified the correctness of our implementation of the ice shelf model by checking that numerical results converge at the expected rate to an analytical solution for the velocity with a linearly sloping thickness in a rectangular domain \citep{greve2009dynamics}.
When there is basal friction, the shallow stream equations do not have a simple analytical solution.
To get an exact solution, we chose the ice velocity and thickness a priori and used the computer algebra system SymPy \citep{sympy} to generate a friction coefficient that makes these fields an exact solution.
We then checked that numerical results converge to this manufactured solution at the expected rate \citep{roache2002code}.

Several studies have compared the shallow stream approximation to 3D models such as Blatter-Pattyn or full Stokes \citep{pattyn2013grounding}.
The most appreciable difference between lower- and higher-order models occurs near the glacier grounding line, where the full Stokes equations can represent bridging stresses \citep{van2013fundamentals}.
The lack of vertical strain rates in the ice viscosity in 2D models can also lead to different equilibrium grounding line positions under the same external forcing.
The advantage of the SSA is that that it can capture most of the overall flow features of an individual fast-flowing glacier but at much lower computational cost than the Stokes equations.

@@

Each of these techniques has appeared in the literature on glacier modeling before but rarely all in the same place for a 3D model.
\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband models.
\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem, but this model used a finite difference discretization in the vertical direction.
The model used in \citet{brinkerhoff2015dynamics} is the closest to the one we present below.
This work used both terrain-following coordinates and a tensor product basis of Lagrange finite elements in the vertical and higher-degree polynomials in a single vertical layer.
For the vertical basis functions, they used a plug flow mode and one shear mode.

@@

Rather than divide the spatial domain into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
This type of basis, in which different shape functions are used in different dimensions, is called a \emph{tensor product} element \citep{mcrae2016automated}.
Given a set of finite element basis functions $\{\phi_k(x, y)\}$ defined on the 2D domain $\Omega$ and a set of basis functions $\{\psi_l(\zeta)\}$ defined on the unit interval $[0, 1]$, the tensor product finite element basis $\{\Phi_{kl}\}$ on the extruded domain is defined as
\begin{equation}
    \Phi_{kl}(x, y, \zeta) = \phi_k(x, y)\psi_l(\zeta).
\end{equation}
For example, we can use piecewise linear or quadratic elements on triangles and use quintic or higher degree polynomials in the vertical.
Rather than use the usual Lagrange interpolating polynomial basis in the vertical dimension, we can instead use the Legendre polynomial basis.
The Legendre polynomials are mutually orthogonal and this choice makes the mass matrix block-diagonal.
Our main reason for choosing to build icepack using the Firedrake package was because it natively supports tensor product elements.

@@

The user then has to decide how many vertical modes are sufficient.
Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a ``smoke test'' for the hybrid model.
The degree-2 model is the most minimalistic model that still exhibits vertical shear and can satisfy the zero-stress boundary condition at the ice surface.
Going up to a degree-4 model is sufficient to capture the exact solution for the shallow ice approximation.
In the tutorial notebooks for icepack, we use up to degrees 2 and 4, but the test suite checks up to degree 8.

\citet{brinkerhoff2015dynamics} used a similar approach to the one described above, with one vertical basis function for plug flow and one for shear flow.
The shear flow basis function was chosen to be exact assuming the SIA balance with cold ice, together with a heuristic approximation for polythermal ice.
By contrast, our approach allows for an arbitrary number of shear modes, although using a high-degree basis in the presence of a cold-temperate transition surface will introduce ringing artifacts.
In principle, one could confront this problem by using two layers -- one cold and one temperate -- instead of a single layer.
The extruded mesh capabilities in Firedrake provide the needed functionality for implementing such a model.
We have not yet implemented a multilayer model but this will be the subject of future work.

@@

The heat transport model assumes that the user will supply a volumetric heating rate, but the model itself does not calculate the heating rate from other fields.
The simplest description for the shear heating rate is
\begin{equation}
    q = \tau : \dot\varepsilon = A^{\frac{-1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}.
\end{equation}
The fluidity factor $A$ in Glen's law is roughly a known function of both temperature and melt fraction and we have included this function in the package.
Having some description of heat transport is thus necessary for describing realistic glacier flows, but it is not sufficient by itself.
While the dependence of fluidity on temperature is known fairly well from laboratory experiments, the dependence on melt fraction is known with much less certainty \citep{cuffey2010physics}.
It is likely that users of this module will want to substitute in their own parameterization for melt fraction dependence.
Other processes such as damage, fabric development, and impurities can affect the fluidity as well.
Moreover, there may be other volumetric heat sources that users wish to account for, such as cryo-hydrologic warming \citep{phillips2010cryo}.
These considerations defy any attempt to have one function that calculates the volumetric heating rate from the other state variables.
Instead, we opted for an interface where users calculate this rate themselves, with tutorials and examples that show how to do so for simple scenarios, leaving the freedom to alter the thermal feedbacks as they see fit.
Our general design principle is that icepack will solve differential equations for the various prognostic fields but it's up to users to decide how these fields are coupled.

@@

Other physical fields besides temperature can influence ice fluidity.
At large spatial scales (> 5 km), crevasse fields affect ice flow by reducing the lateral area over which stress can be transmitted.
Modeling individual fractures is not computationally feasible for large-scale simulations.
Instead, we have implemented the phenomenological model described in \citet{albrecht2014fracture}, which is based on the theory of continuum damage mechanics.
This model is defined in the class \texttt{DamageTransport}.
Prognostic damage models can be broken down into three parts: (1) evolve the damage field based on the membrane stress of the glacier, (2) advect the damage field with flow, and (3) feed the damage field back into the fluidity of the glacier.
Changes in fluidity in turn affect the membrane stress; the coupling between bulk damage and ice flow goes both ways.
The model from \citet{albrecht2014fracture} adds sources of damage where the membrane stress exceeds a critical value and sinks of damage when the principal strain rate is less than a critical value.
These kinds of simplified models ignore complications such as inertial or 3D effects.
They are intended more to capture some important features of the real physical system ecnomically than to describe it exactly.

@@

The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
To solve the advection equation equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}, but this scheme introduces a tuneable stabilization parameter.
In the interest of simplicity, we instead default to the implicit Euler scheme, which has no such parameters.
The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.

Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy (CFL) condition.
The use of an unconditionally stable scheme guarantees that they will get an answer, rather than a runtime error, should they try to use a large timestep.
Users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
The extra computational cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
Advanced users who are interested in maximizing performance can subclass the solver to implement a faster explicit scheme.

@@

\subsection{Convex optimization}

@@

Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian \citep{pratt1998gauss}.
Each iteration of Gauss-Newton is more expensive than that of BFGS or gradient descent because it requires the solution of a more complex linear system than just the mass matrix.
The Gauss-Newton method converges fastest by far in virtually every test case we have found, in some instances by up to factor of 50.

@@

One of the main goals for icepack is to create a tool that is accessible to researchers who might not be experts in scientific computing.
Previous work on numerical modeling of glacier flow has focused largely on the technical details of the models themselves -- does a given solver converge with the accuracy expected from finite element theory; does it scale to large numbers of processors; can models accurately predict grounding line retreat, etc.
The person learning to use an unfamiliar software tool is largely absent from the discussion.
For graduate students or other researchers who are not experts in computational physics, the difficulty of learning to use a particular software package may be more of a rate-limiting factor than the speed or efficiency of that software.

@@

We have presented a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
This package advances the state of the art in this field by providing a platform for easily experimenting with the model physics.
In this paper, we have presented three demonstrations of this feature:
\begin{enumerate}
    \item coupling a model of ice shelf flow to a phenomenological model of damage,
    \item changing the sliding law in a simulation of a marine-terminating glacier, and
    \item inferring the fluidity of a floating ice shelf in a way that guarantees positivity.
\end{enumerate}
The phyics of how glaciers interact with their environment are not completely understood.
Consequently, the ability to change components of the model is an essential feature for any tool aimed at researchers in this field.

>>>>

There are four \emph{diagnostic} or \emph{stress balance} models implemented in icepack.
For each of the diagnostic models, we use a formulation of the physics based on a \emph{minimization} or \emph{action principle} \citep{dukowicz2010consistent}.
Action principles are completely equivalent to the usual weak form of a partial differential equation but have certain numerical advantages, as described below.

@@

The action principle is especially useful for designing a robust numerical solver, the implementation of which will be described in \S\ref{subsec:convex-optimization}.
For viscous flow problems near to steady-state, the action is \emph{convex} as a function of the ice velocity, i.e. its second derivative is positive-definite.
Convexity implies that the action functional has a unique minimizer and that, with an appropriate line search strategy, Newton's method will converge from any initial guess.
Algorithms for minimizing convex functionals have better convergence guarantees than algorithms for solving general nonlinear systems of equations while having no additional computational cost \citep{nocedal2006numerical}.
Both formulations are mathematically equivalent.

@@

The shallow ice model can be derived from the Blatter-Pattyn approximation by assuming ice flow is dominated by bed-parallel shear and that surface and basal slopes are small, i.e. $\partial u/\partial x \ll \partial u/\partial z$.
The result is a 2D system of equations for depth-averaged horizontal velocity.
The class in icepack that represents this physics model is called \texttt{ShallowIce}.
The terms in the action functional are:
\begin{align}
    \text{localization} & =\frac{1}{2}\int_\Omega u\cdot u\,dx\\\
    \text{gravity} & = \int_\Omega\frac{2A(\rho_I g)^n}{n+2} h^{n + 1}|\nabla s|^{n - 1}\nabla s\cdot u\,dx\\
    \text{penalty} & = \frac{1}{2}\int_\Omega \ell^2\nabla u\cdot \nabla u\,dx
\end{align}
and the action functional is
\begin{equation}
    J = \text{localization} + \text{gravity} + \text{penalty}.
\end{equation}
The default value of the length scale $\ell$ in the penalty term is defined as:
\begin{equation}
    \ell = 2\max\{\text{cell diameter}, 5h\}
\end{equation}
but users can adjust this to the value of their choice.
This penalty term smooths over numerical artifacts, especially near the ice margins and termini.
In these regions the shallow ice approximation is less applicable, so the error in solving a different set of equations is small compared to the inherent modeling error in using these equations in the first place.

@@

We verified the correctness of our implementation of the ice shelf model by checking that numerical results converge at the expected rate to an analytical solution for the velocity with a linearly sloping thickness in a rectangular domain \citep{greve2009dynamics}.
There are analytical solutions to the shallow stream equations with basal friction \citep{bodhvarsson1955flow, bueler2014exact}; these solutions are more complex algebraically than those for the case without basal friction.
To get an exact solution, we chose the ice velocity and thickness a priori and used the computer algebra system SymPy \citep{sympy} to generate a friction coefficient that makes these fields an exact solution.
We hand-tuned the input parameters so that the values of the manufactured friction coefficient were within reasonable bounds.
We then checked that numerical results converge to this manufactured solution at the expected rate \citep{roache2002code}.

Several studies have compared the shallow stream approximation to 3D models such as Blatter-Pattyn or full Stokes \citep{pattyn2013grounding}.
The most appreciable difference between lower- and higher-order models occurs near the glacier grounding line, where the full Stokes equations can represent bridging stresses \citep{van2013fundamentals}.
The lack of vertical strain rates in the ice viscosity in 2D models can also lead to different equilibrium grounding line positions under the same external forcing.
The advantage of the SSA is that that it can capture most of the overall flow features of fast-flowing glaciers but at much lower computational cost than the Stokes equations.

@@

Each of these techniques has appeared in the literature on glacier modeling before but rarely all in the same place for a 3D model.
\citet{langdon1978numerical} and \cite{bassis2010hamilton} used variational principles and vertical spectral methods, but these works considered only flowband models.
\citet{kleiner2014numerical} used terrain-following coordinates for 3D glacier flow modeling, but they discretized the problem with finite difference methods in every direction and did not take into account the variational formulation of the diagnostic model.
\citet{jouvet2015multilayer} used vertical semi-discretization of the variational problem, but this model used a finite difference discretization in the vertical direction.
The model used in \citet{brinkerhoff2015dynamics} is the closest to the one we present below.
This work used both terrain-following coordinates and a tensor product basis of Lagrange finite elements in the horizontal and higher-degree polynomials in a single vertical layer.
For the vertical basis functions, they used a plug flow mode and one shear mode.

@@

Rather than divide the spatial domain into many vertical layers, we can instead use only one vertical layer and increase the polynomial degree in the vertical direction to obtain higher resolution.
This type of basis, in which different shape functions are used in different dimensions, is called a \emph{tensor product} element \citep{mcrae2016automated}.
Given a set of finite element basis functions $\{\phi_k(x, y)\}$ defined on the 2D domain $\Omega$ and a set of basis functions $\{\psi_l(\zeta)\}$ defined on the unit interval $[0, 1]$, the tensor product finite element basis $\{\Phi_{kl}\}$ on the extruded domain is defined as
\begin{equation}
    \Phi_{kl}(x, y, \zeta) = \phi_k(x, y)\psi_l(\zeta).
\end{equation}
For example, we can use piecewise linear or quadratic elements on (horizontal) triangles and use quintic or higher degree polynomials in the vertical.
Rather than use the usual Lagrange elements in the vertical dimension, we can instead use the Legendre polynomial basis.
The Legendre polynomials are mutually orthogonal, which makes the mass matrix block-diagonal, and provide better approximation properties \citep{szabo2004p}.
Our main reason for choosing to build icepack using the Firedrake package was because it natively supports tensor product elements.

@@

The user then has to decide how many vertical modes are sufficient.
Using only degree 0 is exactly equivalent to the shallow stream approximation and we use this fact as a sanity test for the hybrid model.
The degree-2 model is the minimal model that still exhibits vertical shear and can satisfy the zero-stress boundary condition at the ice surface.
Going up to a degree-4 model is sufficient to capture the exact solution for the shallow ice approximation.
In the tutorial notebooks for icepack, we use up to degrees 2 and 4, but the test suite checks up to degree 8.

\citet{brinkerhoff2015dynamics} used a similar approach to the one described above, with one vertical basis function for plug flow and one for shear flow.
The shear flow basis function was chosen to be exact assuming the SIA balance with cold ice, together with a heuristic approximation for polythermal ice.
By contrast, our approach allows for an arbitrary number of shear modes.

@@

The heat transport model assumes that the user will supply a volumetric heating rate, but the model itself does not calculate the heating rate from other fields.
The simplest description for the shear heating rate is
\begin{equation}
    q = \tau : \dot\varepsilon = A^{\frac{-1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}.
\end{equation}
The fluidity factor $A$ in Glen's law is roughly a known function of both temperature and melt fraction and we have included this function in the package.
Parameterizing fluidity is not sufficient by itself however.
While the dependence of fluidity on temperature is known fairly well from laboratory experiments, the dependence on melt fraction is known with much less certainty \citep{cuffey2010physics}.
Some users of this module may want to substitute in their own parameterization for melt fraction dependence.
Other processes such as damage, fabric development, and impurities can affect the fluidity as well.
Moreover, there may be other volumetric heat sources that users wish to account for, such as cryo-hydrologic warming \citep{phillips2010cryo}.
These considerations defy any attempt to have one function that calculates the volumetric heating rate from the other state variables.
Instead, we opted for an interface where users calculate this rate themselves, with tutorials and examples that show how to do so for simple scenarios, leaving the freedom to alter the thermal feedbacks as they see fit.
Our general design principle is that icepack will solve differential equations for the various prognostic fields but it's up to users to decide how these fields are coupled.

@@

Other physical fields besides temperature can influence ice fluidity.
At large spatial scales (> 5 km), crevasse fields affect ice flow by reducing the lateral area over which stress can be transmitted.
Modeling individual fractures is not computationally feasible for large-scale simulations.
Instead, we have implemented the phenomenological model described in \citet{albrecht2014fracture}, which is based on the theory of continuum damage mechanics.
This model is defined in the class \texttt{DamageTransport}.
Prognostic damage models can be broken down into three parts: (1) evolve the damage field based on the membrane stress of the glacier, (2) advect the damage field with flow, and (3) feed the damage field back into the fluidity of the glacier.
Changes in fluidity in turn affect the membrane stress; the coupling between bulk damage and ice flow goes both ways.
The model from \citet{albrecht2014fracture} adds sources of damage where the membrane stress exceeds a critical value and sinks of damage when the principal strain rate is less than a critical value.

@@

The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
To solve the advection equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}, but this scheme introduces a tuneable stabilization parameter.
In the interest of simplicity, we instead default to the implicit Euler scheme, which has no such parameters.
The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.

Some glaciologists may be unfamiliar with the Courant-Friedrichs-Lewy (CFL) condition.
The use of an unconditionally stable scheme guarantees that they will get an answer, rather than a runtime error, should they try to use a large timestep.
The extra computational cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
Advanced users who are interested in maximizing performance can subclass the solver to implement a faster explicit scheme.

Icepack currently lacks an adaptive timestepping scheme.
Unconditionally stable schemes allow taking long timesteps, but taking very long timesteps will give inaccurate solutions.
At present, users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
Adaptive timestepping will be added in a future release.

@@

\subsection{Convex optimization}\label{subsec:convex-optimization}

@@

Finally, the Gauss-Newton solver defines an approximation to the ``first-order'' part of the objective functional Hessian \citep{pratt1998gauss, habermann2012reconstruction}.
Each iteration of Gauss-Newton is more expensive than that of BFGS or gradient descent because it requires the solution of a more complex linear system than just the mass matrix.
The Gauss-Newton method converges fastest by far in virtually every test case we have found, in some instances by up to factor of 50.

@@

One of the main goals for icepack is to create a tool that is accessible to researchers who might not be experts in scientific computing.
Previous work on numerical modeling of glacier flow has focused largely on the technical details of the models themselves -- does a given solver converge with the accuracy expected from finite element theory; does it scale to large numbers of processors; can models accurately predict grounding line retreat, etc.
For graduate students or other researchers who are not experts in computational physics, the difficulty of learning to use a particular software package may be more of a rate-limiting factor than the speed or efficiency of that software.

@@

We have presented a new software package called \emph{icepack} for modeling the flow of glaciers and ice sheets.
This package advances the state of the art in this field by providing a platform for easily experimenting with the model physics.
In this paper, we have presented three demonstrations of this feature:
\begin{enumerate}
    \item coupling a model of ice shelf flow to a phenomenological model of damage,
    \item changing the sliding law in a simulation of a marine-terminating glacier, and
    \item inferring the fluidity of a floating ice shelf in a way that guarantees positivity.
\end{enumerate}
The physics of how glaciers interact with their environment are not completely understood.
Consequently, the ability to change components of the model is an essential feature for any tool aimed at researchers in this field.
